\begin{abstract}
World models for partially observed environments must imagine
multiple compatible hidden futures and steer between them under
counterfactual actions. JEPA-family world models do this in latent
space, but a vector-valued latent has no internal structure for
carrying the belief over hidden continuations through blind rollout.
We introduce UWM-JEPA, a JEPA world model with a density-matrix
latent on a joint system--environment space and a learned unitary
predictor. The construction preserves the joint-state spectrum
exactly during rollout, so the predictor itself cannot dissipate the
represented uncertainty.

On a hidden-velocity indicator task requiring five-step forward
simulation under a given action sequence with the target observation
masked, UWM-JEPA reaches $0.77$ accuracy and degrades monotonically
as actions are perturbed; a parameter-matched LSTM-JEPA with the
same objective and action head collapses to majority-class accuracy
($0.53$) under every action condition. Under blind rollout, UWM-JEPA
loses fewer than ten points of probe $R^2$ at short horizons while
vector-latent baselines lose forty-one and sixty-eight; both
nevertheless tie on a held-out context probe, locating the
separation in the predictor rather than the encoder. Action
sensitivity itself requires training against counterfactual rather
than teacher-forced targets, a finding that applies beyond the
unitary parameterisation. For JEPA world models to imagine under
partial observability, latent geometry and predictor dynamics
matter, not encoder capacity alone.
\end{abstract}
